Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-08-31 Writing/NatureWater/2ndRoundRevision/manuscript.tex
manuscripts updated and replies to reviewers created to be reviewed

\caption{The impact of water deficit and aquifer thickness on rainfed and irrigated per-area yields of corn and soybean in US High Plains. Irrigated yield responses to water deficits are dissagregated by aquifer thickness level, considering three quantile groups that capture the range of aquifer thickness values in our sample: (i) 9 m to 30.1 m; (ii) 30.1 m to 67.2 m; and (iii) 67.2 m to 215.8 m. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop.}\label{fig:yield-response}
\end{figure}

@@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around 0.90 (90\%) for an average county-level aquifer thickness of 100m to 0.72 (72\%) for an average county-level aquifer thickness of 10m (See Appendix \ref{test-dif-share} for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean although the effect is less pronounced than for corn, potentially due to soybean's lower water requirements and susceptibility to drought relative to corn \citep{zipper2016drought, ruess2022irrigation}.

@@

\caption{The estimated impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean producing counties in the US High Plains (Note: shaded area represents 95\% confidence interval)}\label{fig:ir-share}
\end{figure}

@@

The overall influence of aquifer depletion on drought-related production losses is a function of both per-area irrigated yield and irrigated area changes highlighted in Sections \ref{impact-intensive} and \ref{impact-share}. Combining both responses, Figure \ref{fig:tot-impact} shows the relationship between water deficit and average per-area production (weighted-average of irrigated and rainfed yields, where the weights are production area shares) for corn and soybean at four values of aquifer thickness: 10, 50, 90, and 130 meters. While similar trends are observed as in Figure \ref{fig:yield-response}, the difference in production loss caused by high water deficits among the aquifer thickness levels are much more pronounced due to lower estimated irrigated area share for smaller aquifer thickness values. For example, for an aquifer thickness of 10 m, reductions in average yields occur for smaller water deficits than are required to trigger reductions in per-area irrigated yields alone. Similarly, the overall magnitude of yield reduction is greater for average yields than for per-area irrigated yields. This is because low aquifer thickness values have a greater estimated share of crop area under rainfed production, which magnifies average yield declines with increasing water deficits.

Figure \ref{fig:dif-tot-impact} shows the difference in average per-area productivity for 50 m, 90 m, and 130 m aquifer thickness levels relative to an aquifer thickness of 10 m for corn and soybean. For corn, aquifer thickness does not have a significant impact on productivity when water deficits are below about 350mm. However, beyond this threshold, higher levels of water deficit result in noticeable and statistically significant reductions to average corn production. This reflects the combination of lower irrigated yields under drought conditions, and, more importantly, reduced overall productivity due to smaller share of land under irrigation in areas and years where aquifer thickness is lower. Similar effects are also observed for soybean, although reductions in average production only occur at a higher deficit threshold (greater than around 600mm) and are smaller in overall magnitude. For example, for a seasonal water deficit of 950mm (average water deficit in 2012, the most extreme drought in our historical record), average per-area corn and soybean productivity are about 25\% and 15\% lower, respectively, when differentiating between areas with the 130m and 10m aquifer thicknesses.

Figure \ref{fig:tot-impact} and Figure \ref{fig:dif-tot-impact} also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950mm, for example, reducing from 130m aquifer thickness to 90m (decline of 40m) would result in only 0.11 tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \ref{fig:ir-share}. Reducing from 90m aquifer thickness to 50m (decline of 40m) would result in 0.77 tonne/ha of production loss for corn, with a further drop from 50m aquifer thickness to 10m (decline of 40m) leading in 2.09 tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \ref{fig:ir-share}), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers ability to effectively and reliably buffer crops against water deficits during the growing season.

@@

\caption{Difference in average yield of corn and soybean for aquifer thickness levels of 50 m, 90 m, and 130 m, relative to a baseline aquifer thickness of 10 m. Shaded areas denote 95\% confidence interval, which when greater than zero indicate average yields for the given aquifer thickness level are statistically significantly greater than for an aquifer thickness of 10 m at the 5\% level of significance.}\label{fig:dif-tot-impact}
\end{figure}

@@

These data were combined to form four sets of data for regression analysis: (1) per-acre corn yield (8773 observations), (2) per-acre soybean yield (5977 observations), (3) share of irrigated corn (2686 observations), and (4) share of irrigated soybean (2014 observations).

@@

\begin{equation}
y_{i,t} = \sum_{j=1}^4 \sigma_j C_j + \sum_{j=1}^4 \sum_{k=1}^K \beta_{j,k}\phi_{j,k}(WB_{i,t})\cdot C_j  + \alpha_i + \lambda_t + v_{i,t} \label{eq:eqn1}
\end{equation}

where county and year are indicated by subscripts \(i\) and \(t\), respectively, \(y\) is crop yield, and \(WB\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0\%-33.3\%)\), \([33.3\%-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) represents the category specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient (see Appendix \ref{spline-basis} for further explanations on how spline-based regression works). $K$ is the total number of spline basis functions. $K = 3$ was chosen for this regression as it minimizes BIC. To control for time-invariant heterogeneity across counties and yearly shocks, county fixed effects (\(\alpha_i\)) and year fixed effects (\(\lambda_t\)) are also included and standard error estimation is clustered by county.

@@

\begin{equation}
    log(\frac{s_{i,t}}{1-s_{i,t}}) = \beta_0 + \sum_{k=1}^K \alpha_{k}\phi_{k}(WB_{i}) + \sum_{l=1}^L \beta_{l}\tau_{l}(AT_{i,t}) + X_i + \theta_{s,t} + \mu_t + v_{i,t} \label{eq:eqn2}
\end{equation}

where \(s_{i,t}\) is the share of irrigated production. The impact of average water balance is captured by \(\sum_{k}^K \alpha_{k}\phi_{k}(WB_{i})\), where \(\phi_{k}(\cdot)\) is the \(k\)th spline basis function and \(\alpha_{k}\) is its coefficient. The impact of aquifer thickness is captured in a similar way with \(\tau_{l}(\cdot)\) and \(\beta_{l}\) being the \(l\)th spline basis function and its coefficient, respectively. $K$ and $L$ are the numbers of basis functions for $WB$ and $AT$, respectively. $K = 4$ and $L = 4$ were chosen as they minimize BIC. County-level soil characteristics including sand percentage, silt percentage, and water holding capacity, are represented by \(X_i\). State-year fixed effects and error term are represented by \(\theta_{s,t}\) and \(v_{i,t}\), respectively. Unlike for our regression models of per-area yields, we do not include county fixed effects to avoid eliminating the majority of variation in the aquifer thickness variable. The vast majority of variation in aquifer thickness is cross-sectional (across counties), rather than over time. Including county fixed effects would effectively eliminate the cross-sectional variation in aquifer thickness. Instead, through robustness checks, we demonstrate that our results are consistent when using state fixed effects. A formal discussion of these results and trade-offs associated with the choice of fixed effects is provided in Appendix \ref{county-fe}.

@@

Figure \ref{fig:test-sat-share} presents the difference in the estimated share of irrigated acres at different levels of aquifer thickness relative to the aquifer thickness level of 130m, which represents a high level of aquifer thickness for our study area that exerts no constraints on water supply for irrigation. Grey shaded areas represented the 95\% confidence intervals for the estimated relationships. Where this range exceeds the red zero line, this indicates that the difference in aquifer thickness (relative to 130 m baseline) leads to a statistically significant (at 5\% level) reduction in irrigated production area share.

@@

\caption{Decline in the share of acres that are irrigated as a function of different levels of aquifer thickness relative to a baseline aquifer thickness level of 130 m (black line). The grey shaded areas represent the 95\% confidence interval for this estimated relationship.}\label{fig:test-sat-share}
\end{figure}

@@

Figure \ref{fig:state-fe-with-without} compares the impact of aquifer thickness on the share of irrigated acres for corn and soybean with different sets of fixed effects. For both corn and soybean, the negative impact of reduction in aquifer thickness on the share of irrigated acres are virtually the same across the sets of fixed effects. Importantly, adding state-year fixed effects, which control for unobservables at the state-year level, does not alter the estimated impacts of aquifer thickness on share of acres irrigated. This indicates that, even though there may be unobservables that are spatially correlated with aquifer thickness beyond soil characteristics, their correlation is unlikely to cause significant bias to our estimation and findings.

>>>>

\caption{The impact of water deficit and aquifer thickness on rainfed and irrigated per-area yields of corn and soybean in US High Plains. Irrigated yield responses to water deficits are dissagregated by aquifer thickness level, considering three quantile groups that capture the range of aquifer thickness values in our sample: (i) 9 m to 30.1 m; (ii) 30.1 m to 67.2 m; and (iii) 67.2 m to 215.8 m. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop. Total number of observations are $8,773$ and $5,977$ for corn and soybean regressions, respectively.}\label{fig:yield-response}
\end{figure}

@@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around 0.90 (90\%) for an average county-level aquifer thickness of 100 m to 0.72 (72\%) for an average county-level aquifer thickness of 10 m (See Appendix \ref{test-dif-share} for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean.

@@

\caption{The estimated impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean producing counties in the US High Plains (Note: shaded area represents 95\% confidence interval). The histograms on the top of each panel show the distribution of county-level aquifer thickness values included for each crop. Total number of observations are $2,523$ and $1,558$ for corn and soybean regressions, respectively.}\label{fig:ir-share}
\end{figure}

@@

The overall influence of aquifer depletion on drought-related production losses is a function of both per-area irrigated yield and irrigated area changes highlighted in Sections \ref{impact-intensive} and \ref{impact-share}. Combining both responses, Figure \ref{fig:tot-impact} shows the relationship between water deficit and average per-area production (weighted-average of irrigated and rainfed yields, where the weights are production area shares) for corn and soybean at four values of aquifer thickness: 10, 40, 70, and 100 meters. While similar trends are observed as in Figure \ref{fig:yield-response}, the difference in production loss caused by high water deficits among the aquifer thickness levels are much more pronounced due to lower estimated irrigated area share for smaller aquifer thickness values. For example, for an aquifer thickness of 10 m, reductions in average yields occur for smaller water deficits than are required to trigger reductions in per-area irrigated yields alone. Similarly, the overall magnitude of yield reduction is greater for average yields than for per-area irrigated yields. This is because low aquifer thickness values have a greater estimated share of crop area under rainfed production, which magnifies average yield declines with increasing water deficits.

Figure \ref{fig:dif-tot-impact} shows the difference in average per-area productivity for 40 m, 70 m, and 100 m aquifer thickness levels relative to an aquifer thickness of 10 m for corn and soybean. For corn, aquifer thickness does not have a significant impact on productivity when water deficits are below about 400 mm. However, beyond this threshold, higher levels of water deficit result in noticeable and statistically significant reductions to average corn production. This reflects the combination of lower irrigated yields under drought conditions, and, more importantly, reduced overall productivity due to smaller share of land under irrigation in areas where aquifer thickness is lower. Similar effects are also observed for soybean, although reductions in average production only occur at a higher deficit threshold (greater than around 600 mm) and are smaller in overall magnitude. For example, for a seasonal water deficit of 950 mm (average water deficit in 2012, the most extreme drought in our historical record), average per-area corn and soybean productivity are about 25\% and 15\% lower, respectively, when differentiating between areas with the 100 m and 10 m aquifer thicknesses.

Figure \ref{fig:tot-impact} and Figure \ref{fig:dif-tot-impact} also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950 mm, for example, reducing from 100 m aquifer thickness to 70 m (decline of 30 m) would result in only 0.09 tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \ref{fig:ir-share}. Reducing from 70 m aquifer thickness to 40 m (decline of 30 m) would result in 0.95 tonne/ha of production loss for corn, with a further drop from 40 m aquifer thickness to 10 m (decline of 30 m) leading in 1.57 tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \ref{fig:ir-share}), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers ability to effectively and reliably buffer crops against water deficits during the growing season.

@@

\caption{Difference in average yield of corn and soybean for aquifer thickness levels of 40 m, 70 m, and 100 m, relative to a baseline aquifer thickness of 10 m. Shaded areas denote 95\% confidence interval, which when greater than zero indicate average yields for the given aquifer thickness level are statistically significantly greater than for an aquifer thickness of 10 m at the 5\% level of significance.}\label{fig:dif-tot-impact}
\end{figure}

@@

These data were combined to form four sets of data for regression analysis: (1) per-acre corn yield ($8,773$ observations), (2) per-acre soybean yield ($5,977$ observations), (3) share of irrigated corn ($2,523$ observations), and (4) share of irrigated soybean ($1,558$ observations).

@@

\begin{equation}
Y_{i,t} = \sum_{j=1}^4 \sigma_j C_j + \sum_{j=1}^4 \sum_{k=1}^K \beta_{j,k}\phi_{j,k}(WB_{i,t})\cdot C_j  + \alpha_i + \lambda_t + v_{i,t} \label{eq:eqn1}
\end{equation}

where county and year are indicated by subscripts \(i\) and \(t\), respectively, \(y\) is crop yield, and \(WB\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0\%-33.3\%)\), \([33.3\%-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) representing the category specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient (see Appendix \ref{spline-basis} for further explanations on how spline-based regression works). $K$ is the total number of spline basis functions. $K = 3$ was chosen for this regression as it minimizes BIC. To control for time-invariant heterogeneity across counties and yearly shocks, county fixed effects (\(\alpha_i\)) and year fixed effects (\(\lambda_t\)) are also included and standard error estimation is clustered by county.

@@

\begin{equation}
    log(\frac{s_{i,t}}{1-s_{i,t}}) = \beta_0 + \sum_{k=1}^K \eta_{k}\phi_{k}(WB_{i}) + \sum_{l=1}^L \beta_{l}\tau_{l}(AT_{i,t}) + X_i + \theta_{s,t} + v_{i,t} \label{eq:eqn2}
\end{equation}

where \(s_{i,t}\) is the share of irrigated production. The impact of average water balance is captured by \(\sum_{k}^K \eta_{k}\phi_{k}(WB_{i})\), where \(\phi_{k}(\cdot)\) is the \(k\)th spline basis function and \(\eta_{k}\) is its coefficient. The impact of aquifer thickness is captured in a similar way with \(\tau_{l}(\cdot)\) and \(\beta_{l}\) being the \(l\)th spline basis function and its coefficient, respectively. $K$ and $L$ are the numbers of basis functions for $WB$ and $AT$, respectively. $K = 4$ and $L = 4$ were chosen as they minimize BIC. County-level soil characteristics including sand percentage, silt percentage, and water holding capacity, are represented by \(X_i\). State-year fixed effects and error term are represented by \(\theta_{s,t}\) and \(v_{i,t}\), respectively. Unlike for our regression models of per-area yields, we do not include county fixed effects to avoid eliminating the majority of variation in the aquifer thickness variable. The vast majority of variation in aquifer thickness is cross-sectional (across counties), rather than over time. Including county fixed effects would effectively eliminate the cross-sectional variation in aquifer thickness. A formal discussion of the trade-offs associated with the choice of fixed effects is provided in Appendix \ref{county-fe}.

@@

Figure \ref{fig:test-sat-share} presents the difference in the estimated share of irrigated acres at different levels of aquifer thickness relative to the aquifer thickness level of 100 m, which represents a high level of aquifer thickness for our study area that exerts no constraints on water supply for irrigation. Grey shaded areas represented the 95\% confidence intervals for the estimated relationships. Where this range exceeds the red zero line, this indicates that the difference in aquifer thickness (relative to 100 m baseline) leads to a statistically significant (at 5\% level) reduction in irrigated production area share.

@@

\caption{Decline in the share of acres that are irrigated as a function of different levels of aquifer thickness relative to a baseline aquifer thickness level of 100 m (black line). The grey shaded areas represent the 95\% confidence interval for this estimated relationship.}\label{fig:test-sat-share}
\end{figure}

@@

Figure \ref{fig:state-fe-with-without} compares the impact of aquifer thickness on the share of irrigated acres for corn and soybean with different sets of fixed effects. For both corn and soybean, the negative impact of reduction in aquifer thickness on the share of irrigated acres are virtually the same across the sets of fixed effects. Importantly, adding state-year fixed effects, which control for unobservables at the state-year level, does not alter the estimated impacts of aquifer thickness on share of acres irrigated. This indicates that, even though there may be unobservables that are spatially correlated with aquifer thickness beyond soil characteristics, their correlation may not cause significant bias to our estimation and findings. However, it is not possible to understand the true consequences of not being able to include county-fixed effects cannot be fully investigated with our datasets, and we defer this matter to future studies with less spatially disaggregated data with enough temporal variations in aquifer thickness levels.

### 2023-08-31 Writing/manuscript.Rmd
manuscripts updated and replies to reviewers created to be reviewed

Figure \@ref(fig:tot-impact) and Figure \@ref(fig:dif-tot-impact) also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950mm, for example, reducing from 130m aquifer thickness to 90m (decline of 40m) would result in only `r round(res_at_950[sat == 130 & crop == "corn", avg_yield] - res_at_950[sat == 90 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \@ref(fig:ir-share). Reducing from 90m aquifer thickness to 50m (decline of 40m) would result in `r round(res_at_950[sat == 90 & crop == "corn", avg_yield] - res_at_950[sat == 50 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn, with a further drop from 50m aquifer thickness to 10m (decline of 40m) leading in `r round(res_at_950[sat == 50 & crop == "corn", avg_yield] - res_at_950[sat == 10 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \@ref(fig:ir-share)), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers ability to effectively and reliably buffer crops against water deficits during the growing season.

>>>>

Figure \@ref(fig:tot-impact) and Figure \@ref(fig:dif-tot-impact) also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950mm, for example, reducing from 130m aquifer thickness to 90m (decline of 40m) would result in only `r round(res_at_950[sat == 100 & crop == "corn", avg_yield] - res_at_950[sat == 70 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \@ref(fig:ir-share). Reducing from 90m aquifer thickness to 50m (decline of 40m) would result in `r round(res_at_950[sat == 70 & crop == "corn", avg_yield] - res_at_950[sat == 40 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn, with a further drop from 50m aquifer thickness to 10m (decline of 40m) leading in `r round(res_at_950[sat == 40 & crop == "corn", avg_yield] - res_at_950[sat == 10 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \@ref(fig:ir-share)), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers ability to effectively and reliably buffer crops against water deficits during the growing season.
